---
title: "R_review_part01"
author: "Kristin Henderson"
date: "2025-08-26"
output: html_document
---

This review is based on the book: https://r4ds.hadley.nz/

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("ggthemes")
# install.packages("palmerpenguins")
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
```

## Data Viz Basics

```{r}
penguins
```

```{r}
?penguins
```

Motivating question:
What does the relationship between flipper length and body mass look like?

Create an empty graph
```{r}
ggplot(data = penguins)
```

Define how variables are mapped to the visual properties (aesthetics) of the plot. For now, just map variables to the x and y aesthetics.
```{r}
ggplot(data = penguins, 
       mapping = aes(x = flipper_length_mm, y = body_mass_g))
```

Now articulate how to represent the observations on the plot by defining a geom (geometrical object).
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) + 
  geom_point()
```

When a categorical variable is mapped to an aesthetic, scaling is implemented. ggplot2 automatically assignes a unique value of the aesthetic (here a color) to the variable and by default adds a legend.
```{r warning=FALSE, message=FALSE}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y=body_mass_g, color = species)
) + 
  geom_point()
```

Add a smooth curve that displays the relationship between x and y. When aesthetic mappings are defined in ggplot(), they are assigned at the global level and get passed down to each subsequent geom layer.
```{r warning=FALSE, message=FALSE}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species) 
  ) +
    geom_point() + 
    geom_smooth(method = 'lm')
```

This time add the aesthetic mapping for colored points at the local level.
```{r warning=FALSE, message=FALSE}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species)) + 
  geom_smooth(method = 'lm')
```

Map to a shape aesthetic in addition to color.
```{r warning=FALSE, message=FALSE}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = 'lm')
```

```{r}
library(tswge)
walmart = read.csv(file.choose(), header = TRUE)
Stor8Item1 = walmart %>% filter(item == 1 & store == 8)
head(walmart)
plotts.wge(Stor8Item1$sales)
acf(Stor8Item1$sales)
```

```{r}
#truncation 400
acf(Stor8Item1$sales)
parzen.wge(walmart$sales)
```

